Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2014-11-17 docs/papers/shallow_performance/shallow_perf.tex
Update table of results and add plot of same data

%% shallow_perf.tex
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{./IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

@@

\subsection{The \psykal Approach}
Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the e.g. oceanographic demands.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge &       & 14.0.1.106  & 8.3.3   &          \\
IBM A2 (BGQ)     &       &             &         & 14.1     \\
IBM Power 8      &       &             &         &          \\
\hline
\end{tabular}
\end{table}

@@

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
\hline
CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
\hline
Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
\hline
\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
\hline
\end{tabular}
\end{table*}

>>>>

%% shallow_perf.tex
%%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

@@

\subsection{The \psykal Approach}
Proliferation of lightweight cores is resulting in increased
complexity in the programming models required to achieve performance.
Combine with the complexity of e.g. a model simulating circulation in
the global ocean. Significant challenge in software engineering.
Propose an approach for meeting this challenge by separating the
computational science demands from the {\it e.g.} oceanographic
demands.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM bgxlf\\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM A2 (BGQ)     &       &             &         & 14.1     \\
IBM Power 8      &       &             &         &          \\
\hline
\end{tabular}
\end{table}

@@

In Table~\ref{TABLE_opt_breakdown} we show detailed timings for the
$256^{2}$ case after each of the transformations has been applied to
the code. For the Gnu compiler, the vanilla \psykal version of shallow
is almost 20 times slower than the original. Copying the code for each
of the kernels into the module from which they are called has the most
dramatic effect on the performance of the compiled code. The next most
significant transformation is that of (manually) in-lining the kernel
code into the bodies of the loops from which they are called. This
gets the \psykal version to within 8\% of the performance of the
original.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu    & Intel   & Gnu & Cray & Intel & IBM bgxlf \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & PowerPC A2 \\
\hline
Original                             & 1.49 & 1.41 & 1.54 & 1.23 & 1.57  &        \\
Vanilla \psykal                      & 25.3 & 1.73 & 28.7 & 1.67 & 1.96  &        \\
Explicitly-sized arrays              & 25.3 & 1.72 & 28.6 & 1.39 & 1.93  & 113.74 \\
Kernels copied into module           & 2.20 & 1.73 & 1.84 & 1.42 & 1.93  &  68.28 \\
Loops fused                          & 2.18 & 1.58 & 1.79 & 1.47 & 1.76  &  71.91 \\
In-lined field-copy                  & 2.17 & 1.58 & 1.77 & 1.46 & 1.75  &  83.36 \\
Fused field-copy                     & 2.18 & 1.85 & 1.99 & 1.30 & 2.07  &  72.62 \\
In-line all kernels                  & 1.64 & 1.43 & 1.88 & 1.30 & 1.59  &        \\
Fully fuse 1st loop nest             & 1.62 & ADD  & 1.55 & 1.25 & 1.55  &        \\
Explicit, constant array/loop bounds & 1.61 & ADD  & 1.56 & 1.23 & 1.54  &        \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 1.2 & 0.6 & 0.0 & -2.0 &     \\
\hline
\end{tabular}
\end{table*}

% 0.32/1.41 = 23% and 0.39/1.57 = 25%
Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of shallow is only
23--25\% slower than the original version. The only code
transformation that produced a significant performance improvement for
this case was the fusion of the computational loops. Further
investigation of the slow-down caused by attempting to fuse the
field-copy operation with the preceeding time-smoothing loop revealed
that the compiler was unable to vectorise the loop after it had
in-lined the body of the time-smoothing kernel. Once this in-lining
was done by hand, the compiler was able to vectorise the loop and the
performance of the original version was recovered.

\begin{figure}
\begin{verbatim}
! Time smoothing
DO J=1,N+1
  DO I=1,M+1
    CALL time_smooth_code(i,j,ufld,unew,uold)
!    uold(i,j) = ufld(i,j) + alpha* &
!        (unew(i,j) - 2.*ufld(i,j) + uold(i,j))

    CALL time_smooth_code(i,j,vfld,vnew,vold)
!    vold(i,j) = vfld(i,j) + alpha* &
!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))

    CALL time_smooth_code(i,j,pfld,pnew,pold)
!    pold(i,j) = pfld(i,j) + alpha* &
!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
  END DO
END DO

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
    Vfld(I,J) = VNEW(I,J)
    Pfld(I,J) = PNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve performance of original shallow with the Intel
  compiler.}
\label{FIG_time_smooth_code}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=85mm]{opt_stages_256}
\caption{The absolute performance of the code at each stage of 
optimisation for the $256^{2}$ domain.}
\label{FIG_opt_stages_256}
\end{figure}
